---
title: "Assignment 5: Data Visualization"
author: "Karly Nocera"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A05_DataVisualization.Rmd") prior to submission.

The completed exercise is due on Tuesday, February 23 at 11:59 pm.



## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse 
and cowplot packages. Upload the NTL-LTER processed data files for nutrients 
and chemistry/physics for Peter and Paul Lakes (both the tidy 
[`NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv`] and the gathered
[`NTL-LTER_Lake_Nutrients_PeterPaulGathered_Processed.csv`] versions) and 
the processed data file for the Niwot Ridge litter dataset. 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r}
#1 Read in datasets and packages
getwd()
library(tidyverse)
library(cowplot)
library(lubridate)

Chem_Nurients_PeterPaul_Processed <- read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv", stringsAsFactors = TRUE)

Nutr_PeterPaul_Gathered <- read.csv("./Data/Processed/NTL-LTER_Lake_Nutrients_PeterPaulGathered_Processed.csv", stringsAsFactors = TRUE)

NIWO_Litter_Processed <- read.csv("./Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv", stringsAsFactors = TRUE)

#2 Fix Dates

### with as.Date
Chem_Nurients_PeterPaul_Processed$sampledate <- as.Date(Chem_Nurients_PeterPaul_Processed$sampledate, format = "%Y-%m-%d")

### with Lubridate 
Nutr_PeterPaul_Gathered$sampledate <- ymd(Nutr_PeterPaul_Gathered$sampledate)

NIWO_Litter_Processed$collectDate <- ymd(NIWO_Litter_Processed$collectDate)

```
## Define your theme
3. Build a theme and set it as your default theme.

```{r}
mytheme <- theme_classic(base_size = 10, base_family = "sans") +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")

theme_set(mytheme)

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values.

```{r}

TotalPhos_Phosphate <- 
  ggplot(Chem_Nurients_PeterPaul_Processed, aes(x = po4, y = tp_ug, color = lakename)) +
  geom_point() +
  xlim(0, 55) +
  ylim(0, 180) +
  geom_smooth(method = lm, color = "black") +
  ylab(expression("Total Phosphorus (ug/L)")) +
  xlab(expression("Phosphate (ug/L)")) +
  labs(color="Lake Name")
  
print(TotalPhos_Phosphate)

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned.

```{r}

# set month as factor to prevent continuous x
Chem_Nurients_PeterPaul_Processed$month <- as.factor(Chem_Nurients_PeterPaul_Processed$month)

Temp_plot <- 
  ggplot(Chem_Nurients_PeterPaul_Processed, aes(x = month, y = temperature_C)) +
  geom_boxplot(aes(color = lakename)) +
  ylab(expression("Temperature - C")) +
  theme(legend.position="none")

print(Temp_plot)

TotalP_plot <- 
  ggplot(Chem_Nurients_PeterPaul_Processed, aes(x = month, y = tp_ug)) +
  geom_boxplot(aes(color = lakename)) +
  ylab(expression("Phosphorous - ug/L")) +
  theme(legend.position="none")

print(TotalP_plot)

TotalN_plot <- 
  ggplot(Chem_Nurients_PeterPaul_Processed, aes(x = month, y = tn_ug)) +
  geom_boxplot(aes(color = lakename)) +
  ylab(expression("Nitrogen - ug/L")) +
  theme(legend.justification = "top")

print(TotalN_plot)

### save legend from last plot that still has legend
legend <- get_legend(TotalN_plot)

### remove that legend from plot
TotalN_plot <- TotalN_plot + theme(legend.position="none")

### cowplot with one legend; graph axis align
### note: when add legend, "graphs cannot be aligned unless axis parameter is set. Placing graphs unaligned."
plot_grid(Temp_plot, TotalP_plot, TotalN_plot, legend, align = "hv", nrow = 1, labels = c('Temp', 'Total Phos', 'Total N'), label_size = 10)

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: Paul Lake has a narrower range of observation values across variables of interest, and often lower than Peter Lake findings. Temperature variation is greatest in the middle of the year, total nitrogen is fairly consistent across the year, and total phosphorous is fairly consistent in Paul Lake but increases steadily from May to October.

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r}
#6. 
Litter.plot <-
  ggplot(subset(NIWO_Litter_Processed, functionalGroup == "Needles"),
    aes(x = collectDate, y = dryMass, color = nlcdClass)) +
  geom_point()

print(Litter.plot)

#7

Litter.facet.plot <-
  ggplot(subset(NIWO_Litter_Processed, functionalGroup == "Needles"),
    aes(x = collectDate, y = dryMass)) +
  geom_point() +
  facet_wrap(vars(nlcdClass), nrow = 1) +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1))

print(Litter.facet.plot)

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: Plot 7, the faceted visual, is more effective because can compare classes side by side while maintaining the same y-axis, can see the "big picture" pattern of each.
