---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Karly Nocera"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A06_GLMs.Rmd") prior to submission.

The completed exercise is due on Tuesday, March 2 at 1:00 pm.

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup}
#1
getwd()
library(tidyverse)
library(htmltools)
library(agricolae)
library(lubridate)
library(viridis)

ChemPhys <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv")

class(ChemPhys$sampledate)
ChemPhys$sampledate <- ymd(ChemPhys$sampledate)

#2

mytheme <- theme_classic(base_size = 10, base_family = "sans") +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")

theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: The change in mean lake temperatures recorded during July across all lakes is zero.
Ha: The change in mean lake temperatures recorded during July across all lakes is different than zero.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4

ChemPhys.tidy <-
  ChemPhys %>%
  mutate(month = month(sampledate)) %>%
  filter(month == 7) %>%
  select(lakename, year4, daynum, depth, temperature_C, month) %>%
  drop_na()

### note: when checking data, see that daynum suggests a different day of the year than the sampledate (ex: 2006-07-04 has daynum 159 even though that day is in June. To adjust, wrangled based on a mutated column for month)

#5

tempvdepth <-
  ggplot(ChemPhys.tidy, aes(x = temperature_C, y = depth, color = depth)) +
  geom_point() +
  scale_y_reverse() +
  xlim(0, 35) +
  xlab(expression("Temperature (C)")) +
  geom_smooth(method = lm) +
  scale_color_viridis(direction = -1)
print(tempvdepth)

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: This figure suggests that shallower depths are associated with warmer temperatures and depper depths with colder temperatures. The distribution of points suggest this may be an exponential rather than linear function, as there are diminishing returns as temperature increases and there is a threshold at which deeper depths are not associated with colder temperatures. 


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7

ChemPhys.regression <- lm(data = ChemPhys.tidy, temperature_C ~ depth)
summary(ChemPhys.regression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: This regression indicates a statistically significant negative relationship between temperature and depth, as seen by p-value below 0.05 for the coefficient with degrees of freedom (392). That is to say, as depth increases every 1m, temperature is predicted to drop -1.9 degrees. The variability in temperature explained by the change in depth is given by the R2 value, at 74.4%.

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9

TPAIC <- lm(data = ChemPhys.tidy, temperature_C ~ year4 + daynum + depth)

step(TPAIC) # identifies daynum and depth as optimal variables

#10

TPmodel <- lm(data = ChemPhys.tidy, temperature_C ~ daynum + depth)
summary(TPmodel)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The final set of explanatory variables are daynum and depth to predict temperature in the multiple regression. This model explains 77.9% of the observed variance, which is an improvement from 74.4% in the single regression model.

---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12

## ANOVA model (aov)

ChemPhys.anova <- aov(data=ChemPhys.tidy, temperature_C ~ lakename)
summary(ChemPhys.anova)

## linear model (lm)

chemPhys.lm.aov <- lm(data = ChemPhys.tidy, temperature_C ~ lakename)
summary(chemPhys.lm.aov)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Cannot reject the null hypothesis that there is a statistically significant difference in mean temperature among the lakes because the p-value for the ANOVA is above 0.05 (at 0.357).

14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty.

```{r scatterplot.2}
#14.

tempvdepth.lake <-
  ggplot(ChemPhys.tidy, aes(x = temperature_C, y = depth, color = lakename)) +
  geom_point(alpha = 0.5) + #transparency
  geom_smooth(method = lm, se = FALSE) +
  xlim(0, 35) + #x is degree, not y
  scale_y_reverse() +
  facet_wrap(vars(lakename), nrow = 1) +
  theme(legend.position="none") +
  xlab(expression("Temperature (C)")) +
  ylab(expression("Depth (m)"))
print(tempvdepth.lake)

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15

TukeyHSD(ChemPhys.anova)

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Two pairs of lakes show similar mean temperatures: 1) Tuesday Lake and East Long Lake; 2) Peter Lake and Paul Lake. However, neither have a p-value less than 0.05 so these results are not statistically significant. In fact, none of the lakes have a mean temperature that is statistically distinct from all the other lakes, because all p-values are above 0.05.

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: The HSD Test can help see if there are any pair-wise relationships and further explore if Peter Lake and Paul lake have distinct mean temperatures.

